Stage-Splitting a Modal Language

### 2014-07-01 paper/semantics.tex
start work on metatheory

% TODO move these to macros.tex once they're finalized
\newcommand{\pval}{\ensuremath{\mathsf{pval}}}
\newcommand{\res}{\ensuremath{\mathsf{res}}}
\renewcommand{\val}{\ensuremath{\mathsf{val}}}

The algorithm described above creates three different kinds of expressions which
cannot be evaluated further at a particular stage:
\begin{itemize}
\item 
Partial values ($\pval$s) are stage \bbone\ terms that have been fully evaluated, 
but which may contain hatted variables. In the example above, 
\verb|(|$\hat {\tt y}$\verb|,7)|.
\item Residuals ($\res$es) are stage \bbtwo\ terms whose stage \bbone\
subexpressions have all been fully evaluated. In the example above,
\verb|(1+2)| and \verb|(let z = y*y*7 in z)|.
\item Values ($\val$s) are stage \bbtwo\ terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|63|.
\end{itemize}
The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future environment} $\xi$ and a partial value $v$.
The future environment is a mapping from newly-created variables 
(which then may appear---hatted---in $v$)
to residuals.  It is the formal manifestation of the floating substitutions in our example.
For all of the normal features of \lang (\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and effect on values as standard evaluation,
and the final future environment is gotten by merging the future environments of the subterms.

@@

% local macros; put them in macros.tex once they are more finalized
\newcommand{\wf}{\ \mathsf{wf}}
\newcommand{\valoo}{\ \bbone\textsf{-}\mathsf{val} @ \bbone}
\newcommand{\valot}{\ \bbone\textsf{-}\mathsf{val} @ \bbtwo}
\newcommand{\valtt}{\ \bbtwo\textsf{-}\mathsf{val} @ \bbtwo}

@@

\begin{definition}
An environment $\xi$ is well-formed ($\xi\wf$) if either:
\begin{enumerate}
\item $\xi = \cdot$; or
\item $\xi = \xi',x\mapsto e$ where
$\Xi,\dom{\xi'};\Gamma\vdash e : B @ \bbtwo$ and
$\Xi,\dom{\xi'};\Gamma\vdash e \valot$.
\end{enumerate}
\end{definition}

@@

\begin{theorem}
If $\diaonesub$ then:
\begin{enumerate}
\item $\typesone e A$ for some 1-type $A$;
\item $\xi\wf$;
\item $\Xi,\dom\xi;\cdot\vdash v : A @ \bbone$; and
\item $\Xi,\dom\xi;\cdot\vdash v\valoo$.
\end{enumerate}
\end{theorem}

\begin{theorem}
If $\diatwosub$ then:
\begin{enumerate}
\item $\typestwo e A$ for some 2-type $A$;
\item $\Xi;\Gamma\vdash q : A @ \bbtwo$; and
\item $\Xi;\Gamma\vdash q\valtt$.
\end{enumerate}
\end{theorem}

>>>>

The algorithm described above creates three different kinds of expressions which
cannot be evaluated further at a particular stage:
\begin{itemize}
\item 
Partial values ($\pvalsym$s) are stage \bbone\ terms that have been fully evaluated, 
but which may contain hatted variables. In the example above, 
\verb|(|$\hat {\tt y}$\verb|,7)|.
\item Residuals ($\ressym$es) are stage \bbtwo\ terms whose stage \bbone\
subexpressions have all been fully evaluated. In the example above,
\verb|(1+2)| and \verb|(let z = y*y*7 in z)|.
\item Values ($\valsym$s) are stage \bbtwo\ terms which are fully evaluated; these
are the results of a computation after both stages have been completed. In the
example above, the term evaluates to \verb|63|.
\end{itemize}
The $\redonesym$ judgment takes an open stage \bbone\ term to a {\em future environment} $\xi$ and a partial value $v$.
The future environment is a mapping from newly-created variables 
(which then may appear---hatted---in $v$)
to residuals.  It is the formal manifestation of the floating substitutions in our example.
For all of the normal features of \lang (\ref{fig:diaSemanticsCore}), first stage evaluation has the same behavior and effect on values as standard evaluation,
and the final future environment is gotten by merging the future environments of the subterms.

@@

@@

\begin{definition}
An environment $\xi$ is well-formed ($\xi\wf$) if either:
\begin{enumerate}
\item $\xi = \cdot$; or
\item $\xi = \xi',x\mapsto e$ where
$\xi'\wf$,
$\Xi,\dom{\xi'};\Gamma\vdash \coltwo{e}{B}$, and
$\Xi,\dom{\xi'};\Gamma\vdash e \res$.
\end{enumerate}
\end{definition}

@@

\begin{theorem}
If $\diaonesub$ and $\typesone e A$ then
\begin{enumerate}
\item $\xi\wf$;
\item $\Xi,\dom\xi;\cdot\vdash \colone{v}{A}$; and
\item $\Xi,\dom\xi;\cdot\vdash v\pval$.
\end{enumerate}
\end{theorem}

\begin{theorem}
If $\diatwosub$ and $\typestwo e A$ then
\begin{enumerate}
\item $\Xi;\Gamma\vdash \coltwo{q}{A}$; and
\item $\Xi;\Gamma\vdash q\val$.
\end{enumerate}
\end{theorem}
